---
title: "Project2_Data_Preparation"
author: "Jason Brown"
date: "2023-11-07"
output: html_document
---

## Project 2 Data Preparation

#### Install and implement libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
pkgs <- sort(c('tidyverse', 'factoextra', 'dbscan', 'cluster', 'mclust', 
  'kernlab', 'e1071', 'scatterpie', 'fpc', 'seriation', 'mlbench', 'GGally'
))
  
pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```


```{r}
pkgs <- sort(c('tidyverse', 'GGally', 'ggcorrplot', 
    'plotly', 'factoextra', 'arules', 'seriation', 
    'sampling', 'caret', 'proxy'))

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```





```{r}

library(tidyverse)
library("factoextra")
library(tidyr)
```



## Import Data


```{r}
selected_cases <- read.csv("selected_features.csv", sep = "")
summary(selected_cases)
```

#### Normalize Data


```{r}
selected_cases <- selected_cases %>% mutate_if(is.character, factor)
dim(selected_cases)
```


```{r}
cases_tibble <- read_csv("Selected_features_tibble.csv")
cases_tibble
```

#### Review Selected Features

```{r}
summary(selected_cases)
```

#### List Selected Features


```{r}
colnames = names(selected_cases)
colnames_sorted <- sort(colnames)

for (i in colnames_sorted)
  print(i)
  features <- i

for ( i in features)
  print(i)
```
#### Create subsets 

```{r}
selected_cases_per_1000 <- selected_cases %>% mutate(white_pop_per_1000 = white_pop/total_pop*1000, black_pop_per_1000 = black_pop/total_pop*1000, hispanic_pop_per_1000 = hispanic_pop/total_pop*1000, asian_pop_per_1000 = asian_pop/total_pop*1000, amerindian_pop_per_1000 = amerindian_pop/total_pop*1000, children_per_1000 = children/total_pop*1000, amerindian_pop_per_1000 = amerindian_pop/total_pop*1000, poverty_per_1000 = poverty/total_pop * 1000, local_commuters_per_1000 = local_commuters/total_pop*1000, commuters_in_groups_per_1000 = commuters_in_groups/total_pop*1000, out_of_town_commuters_per_1000 = out_of_town_commuters/total_pop*1000, two_parent_families_with_young_children_per_1000 = two_parent_families_with_young_children/total_pop*1000, unemployed_pop_per_1000 = unemployed_pop/total_pop*1000, some_college_and_associates_degree_per_1000 = some_college_and_associates_degree/total_pop*1000, pop_determined_poverty_status_per_1000 = pop_determined_poverty_status/total_pop*1000, not_us_citizen_pop_per_1000 = not_us_citizen_pop/total_pop * 1000, masters_degree_per_1000 = masters_degree/total_pop*1000, bachelors_degree_per_1000 = bachelors_degree/total_pop *1000, associates_degree_per_1000 = masters_degree/total_pop *1000, k1t2_students_per_1000 = k12_students/total_pop*1000, income_under_40_per_1000 = income_under_40/total_pop*1000, income_over_40_under_100_per_1000 = income_over_40_under_100/total_pop*1000, income_over_100_per_1000 = income_over_100/total_pop*1000, income_200000_or_more_per_1000 = income_200000_or_more/total_pop*1000, children_in_single_female_hh_per_1000 = children_in_single_female_hh/total_pop*1000, households_public_asst_or_food_stamps_per_1000 = households_public_asst_or_food_stamps/total_pop*1000, median_income_per_1000 = median_income/total_pop*1000, employed_pop_per_1000 = employed_pop/total_pop*1000, total_pop_per_1000 = total_pop)

 
```



#### Remove NaNs

```{r}
colnames = names(selected_cases)
#colnames_sorted <- sort(colnames)

for (i in colnames_sorted)
 selected_cases_per_1000$i <- NULL
#selected_cases_per_1000[,!(names(selected_cases_per_1000) %in% colnames)]
selected_cases_per_1000
```




#### Remove features not per 1000


```{r}
selected_cases_per_1000 = subset(selected_cases_per_1000, select = -c(children, poverty, local_commuters, commuters_in_groups, out_of_town_commuters, two_parent_families_with_young_children, unemployed_pop, some_college_and_associates_degree, pop_determined_poverty_status, not_us_citizen_pop, masters_degree, bachelors_degree, associates_degree, k12_students, income_under_40, income_over_40_under_100, income_over_100, income_200000_or_more, children_in_single_female_hh, households_public_asst_or_food_stamps, employed_pop, total_pop, median_income,
                                                                     
                                                                      confirmed_cases, deaths))
selected_cases_per_1000  
```






## Summarize selected features


```{r pressure, echo=FALSE}
summary(selected_cases)
```


#### Find counties based of majority and race


```{r}
audit_cases <- as.data.frame(selected_cases_per_1000)
```







```{r}
counties_with_maj_black_pop <- audit_cases%>%filter(audit_cases['black_pop_per_1000'] > audit_cases['white_pop_per_1000'], audit_cases['black_pop_per_1000'] > audit_cases['hispanic_pop_per_1000'], audit_cases['black_pop_per_1000'] > audit_cases['amerindian_pop_per_1000'], audit_cases['black_pop_per_1000'] > audit_cases['asian_pop_per_1000'])

counties_with_maj_hispanic_pop <- audit_cases%>%filter(audit_cases['hispanic_pop_per_1000'] > audit_cases['white_pop_per_1000'], audit_cases['hispanic_pop_per_1000'] > audit_cases['black_pop_per_1000'], audit_cases['hispanic_pop_per_1000'] > audit_cases['amerindian_pop_per_1000'], audit_cases['hispanic_pop_per_1000'] > audit_cases['asian_pop_per_1000'])

counties_with_maj_white_pop <- audit_cases%>%filter(audit_cases['white_pop_per_1000'] > audit_cases['black_pop_per_1000'], audit_cases['white_pop_per_1000'] > audit_cases['hispanic_pop_per_1000'],audit_cases['white_pop_per_1000'] > audit_cases['amerindian_pop_per_1000'], audit_cases['white_pop_per_1000'] > audit_cases['asian_pop_per_1000'])

counties_with_maj_amerindian_pop <- audit_cases%>%filter(audit_cases['amerindian_pop_per_1000'] > audit_cases['black_pop_per_1000'], audit_cases['amerindian_pop_per_1000'] > audit_cases['white_pop_per_1000'], audit_cases['amerindian_pop_per_1000'] > audit_cases['hispanic_pop_per_1000'], audit_cases['amerindian_pop_per_1000'] > audit_cases['asian_pop_per_1000'])

counties_with_maj_asian_pop <- audit_cases%>%filter(audit_cases['asian_pop_per_1000'] > audit_cases['black_pop_per_1000'],
audit_cases['asian_pop_per_1000'] > audit_cases['white_pop_per_1000'], audit_cases['asian_pop_per_1000'] > audit_cases['hispanic_pop_per_1000'], audit_cases['asian_pop_per_1000'] > audit_cases['amerindian_pop_per_1000'])
```




#### Check results

```{r}
count(counties_with_maj_asian_pop, 'county_name')
counties_with_maj_asian_pop
```



#### Remove Unneeded features


```{r}
counties_with_maj_hispanic_pop = subset(counties_with_maj_hispanic_pop, select = -c(black_pop, white_pop, hispanic_pop, asian_pop, amerindian_pop))
counties_with_maj_white_pop = subset(counties_with_maj_white_pop, select = -c(black_pop, white_pop, hispanic_pop, asian_pop, amerindian_pop))
counties_with_maj_amerindian_pop = subset(counties_with_maj_amerindian_pop, select = -c(black_pop, white_pop, hispanic_pop, asian_pop, amerindian_pop))
counties_with_maj_black_pop = subset(counties_with_maj_black_pop, select = -c(black_pop, white_pop, hispanic_pop, asian_pop, amerindian_pop))
counties_with_maj_asian_pop = subset(counties_with_maj_asian_pop, select = -c(black_pop, white_pop, hispanic_pop, asian_pop, amerindian_pop))
```


#### Summarize Selected Features Dataset

```{r}
selected_cases |>
  group_by(state) |>
  summarize(across(everything(), mean))
```

#### Statistic Summary for Majority Black Population plus tagging to track counties by race


```{r}
for(i in counties_with_maj_black_pop)
  counties_with_maj_black_pop['race'] = 1
summary(counties_with_maj_black_pop)


```



```{r}
counties_with_maj_black_pop
```



#### Display Majority Black County Count

```{r}
count(counties_with_maj_black_pop, 'county_name')

```

#### Statistic Summary for Majority White Population plus tagging to track counties by race

```{r}
for(i in counties_with_maj_white_pop)
  counties_with_maj_white_pop['race'] = 2
summary(counties_with_maj_white_pop)

```


#### Display Majority White County Count

```{r}
count(counties_with_maj_white_pop, 'county_name')

```

#### Statistic Summary for Majority Hispanic Population plus tagging to track counties by race


```{r}
for(i in counties_with_maj_hispanic_pop)
  counties_with_maj_hispanic_pop['race'] = 3
summary(counties_with_maj_hispanic_pop)
```


#### Display Majority Hispanic County Count

```{r}

count(counties_with_maj_hispanic_pop, 'county_name')

```


#### Statistic Summary for Majority American Indian Population plus tagging to track counties by race



```{r}
for(i in counties_with_maj_amerindian_pop)
  counties_with_maj_amerindian_pop['race'] = 4
summary(counties_with_maj_amerindian_pop)

```


#### Display Majority American Indian County Count



```{r}
count(counties_with_maj_amerindian_pop, 'county_name')
```


#### Statistic Summary for Majority Asian Indian Population plus tagging to track counties by race



```{r}
for(i in counties_with_maj_asian_pop)
  counties_with_maj_asian_pop['race'] = 5
summary(counties_with_maj_asian_pop)

```


#### Display Majority Asian County Count



```{r}
count(counties_with_maj_asian_pop, 'county_name')
```


#### Sampling of majority White counties dataset so as to not block out other races


```{r}
counties_with_maj_white_pop_tibble <- tibble(counties_with_maj_white_pop)
library(sampling)
id2 <- strata(counties_with_maj_white_pop, stratanames = "state", 
              size = c(5,3,5,5,5,5,2,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5,5), method = "srswor")
```

```{r}
id2
```

```{r}
for (i in id2['ID_unit'])
  sliced_sample_maj_white_counties <- counties_with_maj_white_pop %>% slice(i)
sliced_sample_maj_white_counties
```


#### Take a look at some of the dimensions for the tables from the race-based sampling


```{r}
dim(counties_with_maj_black_pop)
```


```{r}
dim(counties_with_maj_amerindian_pop)
```


#### Taking samples so that the majority Asian counties are still represented


```{r}

s_counties_with_maj_amerindian_pop <- counties_with_maj_amerindian_pop %>%as.data.frame() %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_amerindian_pop <- head(s_counties_with_maj_amerindian_pop, n=10) %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_amerindian_pop <- s_counties_with_maj_amerindian_pop %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_amerindian_pop
```



```{r}

s_counties_with_maj_hispanic_pop <- counties_with_maj_hispanic_pop %>%as.data.frame() %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_hispanic_pop <- head(counties_with_maj_hispanic_pop, n=40) %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_hispanic_pop <- s_counties_with_maj_hispanic_pop %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_hispanic_pop
```



```{r}
s_counties_with_maj_black_pop <- counties_with_maj_black_pop %>%as.data.frame() %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_black_pop <- head(s_counties_with_maj_black_pop, n=15) %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_black_pop <- s_counties_with_maj_black_pop %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_black_pop

```





```{r}
s_counties_with_maj_white_pop <- counties_with_maj_white_pop %>%as.data.frame() %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_white_pop <- head(s_counties_with_maj_white_pop, n=50) %>% arrange(desc(deaths_per_1000))
s_counties_with_maj_white_pop <- s_counties_with_maj_white_pop %>% arrange(desc(deaths_per_1000))

s_counties_with_maj_white_pop <- s_counties_with_maj_white_pop %>% filter(!county_name=="Gove County")


s_counties_with_maj_white_pop


```


#### Creating the an aggregate table set of all the sampled data by race


```{r}
combined_classes_on_race <- list(s_counties_with_maj_amerindian_pop, s_counties_with_maj_black_pop, s_counties_with_maj_hispanic_pop, s_counties_with_maj_black_pop, s_counties_with_maj_white_pop, counties_with_maj_asian_pop)
pca_cases <- Reduce(function(x, y) merge(x, y, all=TRUE), combined_classes_on_race) 
pca_cases
#combined_classes_on_race

```


#### Summarize the results and group by race


```{r}
pca_cases |>
  group_by(race)|>
  summarize(across(everything(),mean))
```


#### Principal Component Analysis of the aggregated set


```{r}
plotly::plot_ly(pca_cases, x = ~deaths_per_1000, y = ~death_per_case, z = ~cases_per_1000, 
      color = ~race, size = 1) |> 
  plotly::add_markers()
```






#### Remove nominal features to not skew the data


```{r}
#pca_cases = subset(pca_cases, select = -c(state, county_name))
pca_cases$county_name <- unclass(pca_cases$county_name)

```

```{r}
pca_cases$state <- unclass(pca_cases$state)
```



#### Remove NaNs

```{r}
pca_cases[is.na(pca_cases)] <- 0
pca_cases
```



#### Scale the data



```{r}
#pca_cases <- pca_cases %>% mutate_all(~(scale(.) %>% as.vector))
#pca_cases
scale_numeric <- function(x) 
  x |> 
  mutate(across(where(is.numeric), ~as.vector(scale(.))))

pca_cases.scaled <- pca_cases |> 
  scale_numeric()
pca_cases <- tibble(pca_cases.scaled)
#pca_cases <- tibble(pca_cases.scaled)
```





```{r}
pca_cases 
```




#### Check for outliers

```{r}
# # 
# # # create detect outlier function
detect_outlier <- function(x) {

  # calculate first quantile
  Quantile1 <- quantile(x, probs=.25)

  # calculate third quantile
  Quantile3 <- quantile(x, probs=.75)

  # calculate inter quartile range
  IQR = Quantile3-Quantile1

  # return true or false
  x > Quantile3 + (IQR*1.5) | x < Quantile1 - (IQR*1.5)
}

# create remove outlier function
remove_outlier <- function(dataframe,
                            columns=names(dataframe)) {

  # for loop to traverse in columns vector
  for (col in columns) {

    # remove observation if it satisfies outlier function
    dataframe <- dataframe[!detect_outlier(dataframe[[col]]), ]
  }

  # return dataframe
  print("Remove outliers")
  print(dataframe)
}

audit_cases <- remove_outlier(pca_cases, columns=names(pca_cases))
```



```{r}
 audit_cases$id <- rownames(audit_cases)
  for (i in audit_cases['id'])
    pca_cases_outliers <- pca_cases.scaled[i,]
  for( i in audit_cases['id'])
    pca_cases <- pca_cases %>% filter(!rownames(pca_cases) %in% c(i))
  pca_cases
```



```{r}
#pca_cases <- pca_cases[-c(16,21,24),]
``` 

```{r}
library(dbscan)
library("GGally")
lof <- lof(pca_cases, minPts= 10)
lof
```

```{r}
ggplot(tibble(index = seq_len(length(lof)), lof = sort(lof)), aes(index, lof)) +
  geom_line() +
  geom_hline(yintercept = 1, color = "red", linetype = 2) +
  geom_hline(yintercept = 2.5, color = "blue", linetype = 2)
```

```{r}
ggplot(pca_cases |> add_column(outlier = lof >= 3), aes(race, death_per_case, color = outlier)) +
  geom_point()
```


```{r}
pca_cases.scaled_clean <- pca_cases  |> filter(lof < 3)

km <- kmeans(pca_cases.scaled_clean, centers = 5, nstart = 10)
pca_cases.scaled_clean_km <- pca_cases.scaled_clean|>
  add_column(cluster = factor(km$cluster))
centroids <- as_tibble(km$centers, rownames = "cluster")

ggplot(pca_cases.scaled_clean_km, aes(x = race, y = death_per_case, color = cluster)) + geom_point() +
  geom_point(data = centroids, aes(x = race, y = death_per_case, color = cluster), shape = 3, size = 10)
```


```{r}
pca_cases <- pca_cases.scaled_clean_km

```

#### Dimensions of the aggregated samples

```{r}
dim(pca_cases)
```



#### Fine tune the features for PCA



```{r}
#cases = subset(cases, select = -c(children_per_1000, poverty_per_1000, local_commuters_per_1000, commuters_in_groups_per_1000, out_of_town_commuters_per_1000, two_parent_families_with_young_children_per_1000, unemployed_pop_per_1000, some_college_and_associates_degree_per_1000, pop_determined_poverty_status_per_1000, not_us_citizen_pop_per_1000, masters_degree_per_1000, bachelors_degree_per_1000, associates_degree_per_1000, k12_students_per_1000, income_under_40_per_1000, income_over_40_under_100_per_1000, income_over_100_per_1000, income_200000_or_more_per_1000, children_in_single_female_hh_per_1000, households_public_asst_or_food_stamps_per_1000, median_income_per_1000, employed_pop_per_1000, total_pop_per_1000, confirmed_cases_per_1000, deaths_per_1000))


pca_cases = subset(pca_cases, select = -c(children_per_1000, local_commuters_per_1000, commuters_in_groups_per_1000, out_of_town_commuters_per_1000, some_college_and_associates_degree_per_1000, masters_degree_per_1000, bachelors_degree_per_1000, associates_degree_per_1000, income_under_40_per_1000, income_over_40_under_100_per_1000, income_over_100_per_1000, income_200000_or_more_per_1000, unemployed_pop_per_1000,employed_pop_per_1000, households_public_asst_or_food_stamps_per_1000, pop_determined_poverty_status_per_1000, children_in_single_female_hh_per_1000, poverty_per_1000, two_parent_families_with_young_children_per_1000, k1t2_students_per_1000, median_income_per_1000, total_pop_per_1000, state, county_name))

pca_cases     
```



#### Prep the data for PCA

```{r}
pca_cases_tibble <- tibble(pca_cases)
dim(pca_cases_tibble)
```


#### Summary of the created tibble

```{r}
summary(pca_cases_tibble)
```


#### Remove NaNs again


```{r}
#pca_cases_tibble <- pca_cases_tibble[ , unlist(lapply(pca_cases_tibble, is.numeric))]
pca_cases_tibble[is.na(pca_cases_tibble)] <- 0
pca_cases_tibble
```



```{r}
pca_cases_tibble  <- pca_cases_tibble[ , unlist(lapply(pca_cases_tibble, is.numeric))]
```





#### View the PCA rotations. Basically gives a breakdown on how essential a component is.





```{r}

results <- prcomp(pca_cases_tibble, scale=TRUE)
results$rotation
```

#### View our PCA Graph as the features are projected into a lower dimension


```{r}
biplot(results, scale = 0)
```



#### Another look at our PCA Analysis

```{r}
pca <- prcomp(pca_cases_tibble, 
              scale = TRUE)

```

```{r}
pca$rotation
```




```{r}
pca$rotation <- -1*pca$rotation
```




```{r}

fviz_pca_biplot(pca,
                label="var",
                habillage = pca_cases$race)
```


```{r}
# pca_cases = subset(pca_cases, select = -c(children_per_1000, local_commuters_per_1000, commuters_in_groups_per_1000, out_of_town_commuters_per_1000, some_college_and_associates_degree_per_1000, masters_degree_per_1000, bachelors_degree_per_1000, associates_degree_per_1000, income_under_40_per_1000, income_over_40_under_100_per_1000, income_over_100_per_1000, income_200000_or_more_per_1000, unemployed_pop_per_1000,employed_pop_per_1000, households_public_asst_or_food_stamps_per_1000, pop_determined_poverty_status_per_1000, children_in_single_female_hh_per_1000, poverty_per_1000, two_parent_families_with_young_children_per_1000, k1t2_students_per_1000))
```




```{r}
fviz_pca_biplot(pca,
                geom = c("point", "text"),
                col.ind = "cos2",
                col.var = "black",
                gradient.cols = c("blue","green","red"))
```






```{r}
plot(pca, type = "line")
```



```{r}
fviz_pca(pca)
```



```{r}
pca_cases_tibble 
```


```{r}
lof <- lof(pca_cases_tibble, minPts= 10)
lof
```



```{r}
ggplot(pca_cases_tibble |> add_column(lof = lof), aes(race, death_per_case, color = lof)) +
    geom_point() + scale_color_gradient(low = "gray", high = "red")
```



```{r}
ggplot(tibble(index = seq_len(length(lof)), lof = sort(lof)), aes(index, lof)) +
  geom_line() +
  geom_hline(yintercept = 1, color = "red", linetype = 2) +
  geom_hline(yintercept = 2, color = "blue", linetype = 3)

```



```{r}
pca_cases_tibble_clean <- pca_cases_tibble  |> filter(lof < 5)

km <- kmeans(pca_cases_tibble_clean, centers = 5, nstart = 10)
pca_cases_tibble_clean_km <- pca_cases_tibble_clean|>
  add_column(cluster = factor(km$cluster))
centroids <- as_tibble(km$centers, rownames = "cluster")

ggplot(pca_cases_tibble_clean_km, aes(x = race, y = death_per_case, color = cluster)) + geom_point() +
  geom_point(data = centroids, aes(x = race, y = death_per_case, color = cluster), shape = 3, size = 10)
```




#### Remove outliers found in the PCA tibble


```{r}
#pca_cases_tibble <- pca_cases_tibble[-c(1,12,23,34,45,56,67,78),]
```



#### Good clusterabilit above 80



```{r}
pca_cases_tibble <-pca_cases_tibble_clean
get_clust_tendency(pca_cases_tibble, n = 5)
```



#### View the LOF


```{r}
ggplot(tibble(index = seq_len(length(lof)), lof = sort(lof)), aes(index, lof)) +
  geom_line() +
  geom_hline(yintercept = 1, color = "red", linetype = 2) +
  geom_hline(yintercept = 1.5, color = "blue", linetype = 3)
```



```{r}
set.seed(1234)
ks <- 2:10
```


```{r}
WCSS <- sapply(ks, FUN = function(k) {
  kmeans(pca_cases_tibble, centers = k, nstart = 5)$tot.withinss
  })

ggplot(tibble(ks, WCSS), aes(ks, WCSS)) + geom_line() +
  geom_vline(xintercept = 5, color = "red", linetype = 2)
```






#### Implement KMeans Cluster, set number of clusters


```{r}
km <- kmeans(pca_cases_tibble, centers = 5, nstart = 10)
km
```


```{r}
pca_clustered <- pca_cases_tibble|> add_column(cluster = factor(km$cluster))
pca_clustered
```



```{r}
km <- kmeans(pca_clustered, centers = 5, nstart = 10)
km
```




#### Dipslay the final cluster


```{r}
ggplot(pca_clustered, aes(x = death_per_case, y = cases_per_1000, colour = cluster), after_stat()) + geom_point() 

```


```{r}
df_new <- pca_clustered[ , unlist(lapply(pca_clustered, is.numeric))]
```




```{r}
library(factoextra)
fviz_cluster(km, data = df_new, centroids = TRUE, repel = TRUE, ellipse.type = "norm")
```


```{r}
ggplot(pivot_longer(centroids, cols = c(race, death_per_case, deaths_per_1000, cases_per_1000, asian_pop_per_1000, amerindian_pop_per_1000, black_pop_per_1000, hispanic_pop_per_1000, white_pop_per_1000), names_to = "feature"),
  aes(x = value, y = feature, fill = cluster)) +
  geom_bar(stat = "identity") +
  facet_grid(rows = vars(cluster))
```




```{r}
library(cluster)
plot(silhouette(km$cluster, d))
```


```{r}
fviz_silhouette(silhouette(km$cluster, d))
```



```{r}
WCSS <- sapply(ks, FUN = function(k) {
  kmeans(pca_clustered, centers = k, nstart = 5)$tot.withinss
  })

ggplot(tibble(ks, WCSS), aes(ks, WCSS)) + geom_line() +
  geom_vline(xintercept = 5, color = "red", linetype = 2)
```




```{r}
scale_numeric <- function(x) mutate_if(x, is.numeric, function(y) as.vector(scale(y)))

pca_clustered_scaled <- pca_clustered |> 
  scale_numeric()
summary(pca_clustered_scaled)
kNNdistplot(pca_cases.scaled, k = 5)
abline(h = .32, col = "red")
```



```{r}
d <- dist(pca_clustered)
hc <- hclust(d, method = "complete")
plot(hc)
```


```{r}
fviz_dend(hc, k = 5)
```




```{r}
clusters <- cutree(hc, k = 4)
cluster_complete <- pca_clustered 
cluster_complete
```


```{r}
ggplot(cluster_complete, aes(race, death_per_case, color = cluster)) +
  geom_point()
```



```{r}
cluster_complete <- cluster_complete[ , unlist(lapply(cluster_complete, is.numeric))]
fviz_cluster(list(data = cluster_complete, cluster = cutree(hc, k = 5)), geom = "point")

```



```{r}
hc_single <- hclust(d, method = "single")
fviz_dend(hc_single, k = 5)
```


```{r}
fviz_cluster(list(data = cluster_complete, cluster = cutree(hc_single, k = 5)), geom = "point")
```




```{r}
library("kernlab")
```


```{r}
cluster_spec <- specc(as.matrix(cluster_complete), centers = 5)
cluster_spec
```



```{r}
ggplot(cluster_complete |> 
         add_column(cluster = factor(cluster_spec)),
  aes(race, death_per_case, color = cluster)) + 
  geom_point()
```




```{r}
library("e1071")

cluster_cmeans <- cmeans(as.matrix(cluster_complete), centers = 5)
cluster_cmeans
```




```{r}
library("scatterpie")
ggplot()  +
  geom_scatterpie(data = cbind(cluster_complete, cluster_cmeans$membership),
    aes(x = race, y = death_per_case), cols = colnames(cluster_cmeans$membership), legend_name = "Membership") + coord_equal()
```




```{r}
fpc::cluster.stats(d, km$cluster)
```


```{r}
sapply(
  list(
    km = km$cluster,
    hc_compl = cutree(hc, k = 4),
    hc_single = cutree(hc_single, k = 4)
  ),
  FUN = function(x)
    fpc::cluster.stats(d, x))[c("within.cluster.ss", "avg.silwidth"), ]
```




















